---
title: "ab20962_EMATM0061_A"
author: "Noah Sheldon"
date: "19/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Section A: Data Wrangling

The dataset used for data wrangling task is hydrographic data from the California Cooperative Oceanic Fisheries Investigations (CalCOFI). 
Data can be accessed by clicking [here](https://www.calcofi.org/ccdata.html).

### Importing Required Libraries

```{r message= FALSE, warning=FALSE}
#load the library
library(tidyverse)
```
Tidyverse is a collection of packages created for data science. Few examples of the packages are ggplot2, dplyr, tidyr..etc. These packages are used to do analysis. 

Loading a library helps us to use the inbuilt functions in the library with ease. 

### Question 1 - Load the dataset

```{r}
#load the dataset
calcofi_data_original <- read.csv("194903-201911_Bottle.csv")

#Check the dimensions of the dataset
dim(calcofi_data_original)
```
The calcofi dataset is loaded into the environment using "read.csv" command. This dataset has been assigned the variable name "calcofi_data_original". 

The "calcofi_data_original" has 887018 rows and 62 columns. The "dim" command helps us to see the dimensions of the dataframe. 

### Question 2 - Generate a new data frame called “calcofi_data” 

Out of the 62 columns we are going to work on just 6 columns from the dataset. We will take a subset from "calcofi_data_original" data and name it "calcofi_data". 

Subsetting the data helps us in the following ways:  
1. We work with the rows, colunmns only required.   
2. Saves memory.  
3. Increases performance.  
4. Makes debugging easier.  

The 6 columns are the following. Each column name is follwed by a description of the it:  
1.Depth_ID - Unique Identifier for a particalur depth.  
2.Depthm - Depth of Water in meters.  
3.T_degC - Temperature of Water measured in degrees centigrade.  
4.Salnty - Salanity of Water.  
5.Oxy_µmol.Kg - Oxygen level in micro moles per kilogram.  
6.ChlorA - Acetone extracted chlorophyll-a measured fluorometrically.  


The full data dictionary can accessed by clicking [here](https://new.data.calcofi.org/index.php/database/calcofi-database/bottle-field-descriptions)

We are also renaming the colunms to column names which are easy,understood and readable. 

The new column names are as follows:  
1.depth - Depthm  
2.temperature - T_degC  
3.salinity - Salnty  
4.oxygen - Oxy_µmol.Kg  
5.chlorophyll ChlorA  


```{r}
# selecting 6 columns from the calcofi original dataset and renaming 5 columns
calcofi_data <- calcofi_data_original %>%
  select(Depth_ID,Depthm,T_degC,Salnty,'Oxy_µmol.Kg',ChlorA) %>%
  rename(depth = Depthm, temperature = T_degC,salinity = Salnty, oxygen = 'Oxy_µmol.Kg', chlorophyll = ChlorA)

#The first five rows of calcofi_data
head(calcofi_data)

```

### Question 3 - Writing two R functions: One called “get_year” and one called “get_cast_type”

The function called “get_year” takes elements from the “Depth_ID” column and outputs the year number. 

For example, given a string of the form “19-4903CR-HY-060-0930-05400560-0000A-3”, the “get_year” function should output the number 1949.

```{r}
#Function: get_year 
#Input: Depth_ID 
#Output: Year the reading was recorded
get_year <- function(Depth_ID){
  return(paste(substr(Depth_ID,1,2),substr(Depth_ID,4,5),sep =""))
}
```
The function “get_cast_type” takes elements from the “Depth_ID” column and outputs the cast type. 

For example, given a string of the form “19-4903CR-HY-060-0930-05400560-0000A-3”, the “get_cast_type” function should output the character string “HY”.

```{r}
#Function: get_cast_type 
#Input: Depth_ID 
#Output: Cast type
get_cast_type <- function(Depth_ID){
  return(substr(Depth_ID,11,12))
}
```

```{r}
#Appplying fucntion get_year to extract year column and get_cast_type to get the cast_type column
#Also dropping the depth_id column
calcofi_data <- calcofi_data %>%
  mutate(year = get_year(Depth_ID),
         cast_type = get_cast_type(Depth_ID),
  ) %>%
  select(-Depth_ID)

#Displaying first 5 rows of calcofi_data
head(calcofi_data)
  
```
#### Which of these variables is categorical?  
The column cast_type is the only categorical variable with unique values

```{r}
unique(calcofi_data$cast_type)
```

### Question 4 - Adding decade column

Creating a new column "decade". We will extract the decade column from the year column. 

For example, 1959 will be mapped to 1950 and 2002 will be to 2000.

The logic used is to subtract year modulus 10 column from year column.   

For example, the year passed in is 1959.  
The formula is, decade = year - (year %% 10)  
This gives us, decade = 1959 -  (1959 %% 10)  
decade = 1959 - 9  
decade = 1950  

```{r}
#Creating a new column "decade" from the year column
calcofi_data$decade <- as.numeric(calcofi_data$year) - (as.numeric(calcofi_data$year) %% 10)

#Filtering the data to only rows with depth 100m and decade between 1950 and 2000.

calcofi_data <- calcofi_data %>%
      filter(depth == 100, 
             between(decade,1950,2000))

#Shape(dimension) of calcofi_data
dim(calcofi_data)

```
The dataframe has 26622 rows.

### Question 5 - Summary DataFrame

Summary dataframe with the sample mean, median and standard deviation of temperature, oxygen and salanity grouped by decade. 

```{r message= FALSE, warning=FALSE}
#
summary_calcofi <- calcofi_data %>%
  select(temperature, oxygen, salinity, decade) %>%
  group_by(decade) %>%
  summarise(across(everything(),
                   list(mn = ~mean(.x,na.rm = T),
                        md = ~median(.x,na.rm = T),
                        sd = ~sd(.x,na.rm = T)))) 
```

The function shorten_colnames is a function that takes a column name as argument and returns a shortened column name.

The logic begind this function is that it takes in a string as an argument and outputs concatenated values of the first three characters and the last three characters of the string.

For example, if we pass the column name was temperature_mn. The output of the function will be tem_mn. This tem_mn will be our new column name. 

```{r}
#Function: shorten_colnames 
#Input: col_name 
#Output: shortened col_name
shorten_colnames <- function(col_name){
  return(paste(substr(col_name,1,3),substr(col_name,nchar(col_name)-2,nchar(col_name)), sep = ""))
}

#Applying new shortened col_names to our dataset "summary_calcofi"
names(summary_calcofi)[-1] <- sapply(names(summary_calcofi)[-1], shorten_colnames)

#Printing out the first five rows of the dataset
head(summary_calcofi)
```










